Large and Infinitary Quotient Inductive-Inductive Types

### 2020-01-10 draft/paper.tex
typo
paper

We use cumulativity to reduce bureaucratic overhead when dealing with
constructions at different universe levels. The internal $\Subtype$ is used in
Section \ref{sec:termmodels} to prove cumulativity for general QIIT
algebras. For example, consider natural number algebras at level $i$, given as
the $\Sigma$-type $NatAlg_i :=(Nat : \Set_i)\times Nat\times(Nat \ra
Nat)$. It follows from the subtyping rules that $i \leq j$ implies $NatAlg_{i}
\leq NatAlg_{j}$. However, cumulativity for arbitrary QIIT algebras does not
follow judgmentally; it can only be proven by induction on signatures, hence
the need for $\Subtype$.

@@

\emph{Notational conventions.} We name elements of $\Con$ as $\Gamma$, $\Delta$,
$\Theta$, elements of $\Sub\,\Gamma\,\Delta$ as $\sigma$, $\delta$, $\nu$,
elements of $\Ty\,\Gamma$ as $A$, $B$, $C$, and elements of $\Tm\,\Gamma\,A$ as
$t$, $u$, $v$. CwF components by default support de Bruijn indices, which are
not easily readable. We use instead a nameful notation for binders in context
extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon\,\ext (a : U) \ext (t :
\El\,a))$. We also define a type-theoretic flavor of $\appi$ for convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[\id, u])\\
  & t\appitt u && := (\appi\,t)[\id, u]
\end{alignat*}
We abbreviate non-dependent inductive $\Pii$ as $\blank\arri\blank$, and
likewise we use $\blank\arre\blank$ and $\blank\arrinf\blank$ for non-dependent
external and infinitary functions.

%% We use abbreviations for non-dependent function types:
%% \begin{alignat*}{3}
%%   & A\arri B && := \Pi\,A\,(B[\p])\\
%%   & A \arre\,B &&:= \Pie\,A\,(\lambda\,\_.\,B) \\
%%   & A \arrinf\,B &&:= \Piinf\,A\,(\lambda\,\_.\,B)
%% $A\arri B := \Pi\,A\,(B[\p])$, and $A \arre\,B := \Pie\,A\,(\lambda\,\_.\,B)$, and likewise
%% $A \arrinf\,B := \Piinf\,A\,(\lambda\,\_.\,B)$.

\begin{definition}[Notion of signature]\label{def:signature}
A QIIT signature at level $j$ is a context in an arbitrary $M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_j := (i : \mathbb{N})\ra(M : \ToS_{i,j})\ra \Con_M
\]

@@

\begin{example}{
    Signature for natural numbers. Here, no external types appear, so the level
    can be chosen as $0$.}
\begin{alignat*}{3}
  & \NatSig : \Sig_0 \\
  & \NatSig := \lambda (i :\mathbb{N})(M : \ToS_{i,0}).\\
  & \hspace{1em}(\emptycon_M\,\ext_M\, (N : \U_M) \,\ext_M\,(zero : \El_M\,N)\\
  & \hspace{2.67em}\ext_M (suc : N\arri_M\El_M\,N))
\end{alignat*}
\end{example}

@@

\begin{example}{Infinitary constructors}. The universe $\U$ is closed under
the $\Piinf$ function type, which allows such functions to appear in the domains
of $\Pii$ types. This allows, for example, a signature for trees branching with
arbitrary small types. This is a signature at level 1, since we have $\Set_0$ as
a $\Pie$ domain type.
\begin{alignat*}{3}
& \mathsf{TreeSig} := \\
& \hspace{1em}\emptycon \ext (Tree : \U)\\
& \hspace{1.35em}\ext (node : \Pie\,\Set_0\,(\lambda A.\,(A\,\arrinf\,Tree)\arri\El\,Tree))
\end{alignat*}
\end{example}

%% \begin{example}{Sort equations}. An interesting example is the term lifting
%% rule for cETT (omitting everything except term lifting):
%% \begin{alignat*}{3}
%%   & cETTSig := \\
%%   & ...\,\ext\,Tm{\Lift} : \Pii\,(\Gamma : Con)\,(\Pii\,(A : Ty{\appitt}\Gamma)((\IdU\,(Tm{\appitt}\Gamma{\appitt}A)\,(Tm{\appitt}\Gamma{\appitt}(\Lift_j{\appitt}A))))
%% \end{alignat*}
%% \end{example}

\begin{example}{Recursive equations}. Again, the universe is closed under $\Id$, which allows us to write equations
in $\Pii$ domains. A minimal (and trivial) example:
\begin{alignat*}{3}
  & \mathsf{RecEqSig} :=\\
  & \hspace{0.5em}\emptycon\,\ext (A : \U) \ext (a : \El\,A)\ext (f : \Pii\,(x : A)\,(\Id\,A\,x\,a\arri\El\,A))
\end{alignat*}
More interesting (and complicated) examples for recursive equations are boundary
conditions in various cubical type theories \cite{cohen2016cubical, angiuli2016computational, angiuli2018cartesian}. Note that our $\Id$
allows iterated equations as well, but these are all trivial in the semantics,
where we assume UIP.
\end{example}

@@

\begin{definition}[Notion of induction in an flCwF]\label{def:induction}
Given $\bGamma : \flCwF_i$, we have the following predicate on contexts:
\begin{alignat*}{3}
  & \Inductive : \Con_{\bGamma} \ra \Set_i \\
  & \Inductive\,\Gamma := (A : \Ty_{\bGamma}\,\Gamma)\ra \Tm_{\bGamma}\,\Gamma\,A
\end{alignat*}
\end{definition}

@@

For $\boldsymbol{\sigma \circ \delta}$, the underlying functions are given by
function composition, and the preservation morphisms are given as follows:
\begin{alignat*}{3}
  & (\boldsymbol{\sigma \circ \delta})_{\emptycon}^{-1} :=
    \bsigma\,\bdelta_{\emptycon}^{-1} \circ \bdelta_{\emptycon}^{-1} \\
  & (\boldsymbol{\sigma \circ \delta})_{\ext}^{-1} :=
    \bsigma\,\bdelta_{\ext}^{-1} \circ \bdelta_{\ext}^{-1}
\end{alignat*}

@@

\subsection{Type Substitution}
We aim to define $\boldsymbol{\blank[\blank] : \Ty\,\Delta \ra
  \Sub\,\Gamma\,\Delta \ra \Ty\,\Gamma}$, such that $\boldsymbol{A[\id]} = \bA$
and $\boldsymbol{A[\sigma\circ\delta]} = \boldsymbol{A[\sigma][\delta]}$. The
underlying sets are given by simple composition:
\begin{alignat*}{3}
  & \Con_{\boldsymbol{A[\sigma]}}\,\ulGamma && := \Con_{\bA}\,(\bsigma\,\ulGamma)\\
  & \Sub_{\boldsymbol{A[\sigma]}}\,\Gamma\,\Delta\,\ulsigma && :=
    \Sub_{\bA}\,\Gamma\,\Delta\,(\bsigma\,\ulsigma)\\
  & \Ty_{\boldsymbol{A[\sigma]}}\,\Gamma\,\ulA && :=
      \Ty_{\bA}\,\Gamma\,(\bsigma\,\ulA)\\
  & \Tm_{\boldsymbol{A[\sigma]}}\,\Gamma\,A\,\ult && :=
      \Tm_{\bA}\,\Gamma\,A\,(\bsigma\,\ult)
\end{alignat*}

Moreover, $\id_{\boldsymbol{A[\sigma]}} := \id_{\bA}$,
$\sigma\circ_{\boldsymbol{A[\sigma]}}\delta := \sigma\circ_{\bA}\delta$,
and likewise substitution components are given by corresponding substitution
components in $\bA$.

Context and type formers are given by coercing $\bA$
structures along $\bsigma$ preservation isomorphisms. For example:
\begin{alignat*}{3}
  &\emptycon_{\boldsymbol{A[\sigma]}} && :=
    \coe\,\bsigma_{\emptycon}^{-1}\,\emptycon_{\bA}\\
  &\Gamma\ext_{\boldsymbol{A[\sigma]}}A && :=
    \coe\,\bsigma_{\ext}^{-1}\,(\Gamma\ext_{\bA} A)\\
  &\Id_{\boldsymbol{A[\sigma]}}\,t\,u && :=
    \coe\,\bsigma_{\Id}^{-1}\,(\Id_{\bA}\,t\,u)
\end{alignat*}

Term and substitution formers are given by composing $\coh$-lifted
isomorphisms with term and substitution formers from $\bA$. For example:
\begin{alignat*}{3}
  & \epsilon_{\boldsymbol{A[\sigma]}} && :=
    \coh\,\bsigma_{\emptycon}^{-1}\,\emptycon_{\bA} \circ \epsilon_{\bA}\\
  & \p_{\boldsymbol{A[\sigma]}} && :=
    \p_{\bA} \circ (\coh\,\bsigma_{\ext}^{-1}\,(\Gamma\ext A))^{-1}\\
  & \appK_{\boldsymbol{A[\sigma]}}\,t && :=
    \appK_{\bA}\,((\coh\,\bsigma_{\K}\,(\K\,\Delta))^{-1}\circ t)
\end{alignat*}
Equations for term and type substitution follow from naturality of preservation
isomorphisms in $\bsigma$, $\coe[]$, $\coh[]$ and substitution equations in
$\bA$.

Iso-cleaving is given iso-cleaving in $\bA$ and the action of $\bsigma$ on
isomorphisms, e.g.\ we have $\coe_{\boldsymbol{A[\sigma]}}\,\ulsigma\,\Gamma
:= \coe_{\bA}\,(\bsigma\,\ulsigma)\,\Gamma$.

@@

$\boldsymbol{\blank[\blank] : \Tm\,\Delta\,A \ra (\sigma : \Sub\,\Gamma\,\Delta)
  \ra \Tm\,\Gamma\,(A[\sigma])}$ is given similarly to
$\boldsymbol{\blank\circ\blank}$ in Section \ref{sec:idcomp}. Underlying functions
are given by function composition, and preservation morphisms are also similar:
\begin{alignat*}{3}
  & (\boldsymbol{t[\sigma]})_{\emptycon}^{-1} :=
    \bt\,\bsigma_{\emptycon}^{-1} \circ \bt_{\emptycon}^{-1} \\
  & (\boldsymbol{t[\sigma]})_{\ext}^{-1} :=
    \bt\,\bsigma_{\ext}^{-1} \circ \bt_{\ext}^{-1}
\end{alignat*}

@@

$\boldsymbol{\p : \Sub\,(\Gamma\ext A)\,\Gamma}$ is a strict morphism given by
taking a first projection for each component. $\boldsymbol{\q : \Tm\,(\Gamma\ext
  A)\,(A[\p])}$ is likewise a strict flCwF section given by second projections. Substitution extension $\boldsymbol{\sigma,\,t}$ is given by pointwise combining $\bsigma$ and $\bt$ with metatheoretic $\Sigma$ pairing, e.g.\ $\Con_{\boldsymbol{(\sigma,t)}}\,\ulGamma := (\bsigma\,\ulGamma,\,\bt\,\ulGamma)$.

@@

\begin{definition} For a level $i$, we write $\bSet_i$ for the flCwF of sets where $\Con_{\bSet_i} := \Set_i$ and $\Sub_{\bSet_i}\,\Gamma\,\Delta := \Gamma \ra \Delta$.
\end{definition}

We define $\boldsymbol{\U : \Ty\,\Gamma}$ as the isofibration which is
constantly $\bSet_i$. A constant isofibration does not actually depend on the
base flCwF, and has trivial iso-cleaving where $\coe$-s are identity
functions. Hence, we have $\Con_{\bU}\,\ulGamma := \Set_i$ and
$\Sub_{\bU}\,\Gamma\,\Delta\,\ulsigma := \Gamma \ra \Delta$.

@@

We define $\boldsymbol{\El : \Tm\,\Gamma\,\U \ra \Ty\,\Gamma}$ as discrete
isofibration formation. For $\boldsymbol{a : \Tm\,\Gamma\,\U}$, the underlying
sets of $\bEl\,\ba$ are the following:
\begin{alignat*}{3}
  & \Con_{\bEl\,\ba}\,\ulGamma && := \ba\,\ulGamma\\
  & \Sub_{\bEl\,\ba}\,\Gamma\,\Delta\,\ulsigma && := \ba\,\ulsigma\,\Gamma = \Delta\\
  & \Ty_{\bEl\,\ba}\,\Gamma\,\ulA && := \ba\,\ulA\,\Gamma\\
  & \Tm_{\bEl\,\ba}\,\Gamma\,A\,\ult && := \ba\,\ult\,\Gamma = A
\end{alignat*}
Hence, in $\bEl\,\ba$, $\Sub$ and $\Tm$ are propositional. We use the
isomorphisms $\ba_{\emptycon} : \ba\,\emptycon \simeq \top$ and $\ba_{\ext} :
\ba\,(\ulGamma\ext\ulA) \simeq (\Gamma :
\ba\,\ulGamma)\times(\ba\,\ulA\,\Gamma)$ to define empty context and context
extension:
\begin{alignat*}{3}
  & \emptycon_{\bEl\,\ba} && := \ba_{\emptycon}^{-1}\,\tt\\
  & (\Gamma\ext_{\bEl\,\ba} A) && := \ba_{\ext}^{-1}\,(\Gamma,\,A)
\end{alignat*}
We likewise use preservation isomorphisms to define $\K$, $\Id$ and $\Sigma$.
Context coercion is $\coe\,\ulsigma\,\Gamma := \ba\,\ulsigma\,\Gamma$. Type coercion, for $A : \ba\,\ulA\,\Gamma$ is given as $\coe\,\ult\,A := \ba\,\ult\,(\ba_{\ext}^{-1}\,(\Gamma,\,A))$.

@@

Contexts are products of $\bB$-contexts, and types are products of $\bB$-types,
indexed respectively by contexts and types of $\bEl\,\ba$.
\begin{alignat*}{3}
  & \Con_{(\bPii\,\ba\,\bB)}\,\ulGamma &&:= (\gamma : \ba\,\ulGamma)\ra \Con_{\bB}\,(\ulGamma, \gamma)\\
  & \Ty_{(\bPii\,\ba\,\bB)}\,\Gamma\,\ulA &&:= (\gamma : \ba\,\ulGamma)(a : \ba\,\ulA\,\gamma)\ra \Ty_{\bB}\,(\Gamma\,\gamma)\,(\ulA, a)
\end{alignat*}
Note that since $\bB$ is over the total $\boldsymbol{(\Gamma\ext\El\,a)}$,
$\Con_{\bB}$ has a $\Sigma$-typed argument, and likewise the last argument
of every $\bB$ component.  We could define substitutions similarly, as
products of substitutions:
\begin{alignat*}{3}
  & \Sub_{(\bPii\,\ba\,\bB)}\,\Gamma\,\Delta\,\ulsigma :=
  (\gamma : \ba\,\ulGamma)(\delta : \ba\,\ulDelta)
  (\sigma : \Sub_{(\bEl\,\ba)}\,\gamma\,\delta\,\ulsigma)\\
  &\hspace{6.5em}\ra \Sub_{\bB}\,(\Gamma\,\gamma)\,(\Delta\,\delta)\,(\ulsigma, \sigma)
\end{alignat*}
This would work, but we know that $\Sub_{(\bEl\,\ba)}\,\gamma\,\delta\,\ulsigma$ is defined
as $\ba\,\ulsigma\,\gamma = \delta$, so we can eliminate $\sigma$ by singleton contraction,
and use the following equivalent definition:
\begin{alignat*}{3}
  & \Sub_{(\bPii\,\ba\,\bB)}\,\Gamma\,\Delta\,\ulsigma :=
  (\gamma : \ba\,\ulGamma)\ra\Sub_{\bB}\,(\Gamma\,\gamma)\,(\Delta\,(\ba\,\ulsigma\,\gamma)\,(\ulsigma, \refl)
\end{alignat*}
The benefit of the contracted definition is that it computes preservation laws
in algebra homomorphisms strictly as expected, while the non-contracted
definition computes homomorphisms as functional logical relations.

@@

Iso-cleaving is given by transporting indices backwards in $\bEl\,\ba$ and outputs forwards
in $\bB$:
\begin{alignat*}{3}
  & \coe\,\ulsigma\,\Gamma &&:=
    \lambda\,\gamma.\,\coe_{\bB}\,(\ulsigma,\refl)\,(\Gamma\,(\ba\,(\ulsigma^{-1})\,\gamma))\\
  & \coe\,\ult\,A &&:=
    \lambda\,\gamma\,a.\,\coe_{\bB}\,(\ult,\refl)\,(A\,(\ba\,(\ult^{-1})\,(\ba_{\ext}^{-1}(\gamma,a))))
\end{alignat*}
Likewise, $\coh$-s are given by backwards-forwards $\coh$-s.

@@

The basic idea is that initial algebras can be built from the terms of $\syn_j$.
For example, consider the syntactic signature for natural numbers:
\[
\NatSig := \emptycon\ext(N : \U)\ext(zero : \El\,N)\ext (suc : N
\arri \El\,N)
\]
The type $\Tm_{\syn}\,\NatSig\,(\El_{\syn}\,N)$ is isomorphic to the usual type
of natural numbers, since, intuitively, such terms can only be built from
iterated usage of $zero$ and $suc$. We will build a term algebra for each
signature in this manner. First, we describe the interpretation of $\syn$ in the
$\Set$ model.

>>>>

We use cumulativity to reduce bureaucratic overhead when dealing with
constructions at different universe levels. The internal $\Subtype$ is used in
Section \ref{sec:termmodels} to prove cumulativity for general QIIT
algebras. For example, consider natural number algebras at level $i$, given as
the $\Sigma$-type $NatAlg_i :\equiv(Nat : \Set_i)\times Nat\times(Nat \ra
Nat)$. It follows from the subtyping rules that $i \leq j$ implies $NatAlg_{i}
\leq NatAlg_{j}$. However, cumulativity for arbitrary QIIT algebras does not
follow judgmentally; it can only be proven by induction on signatures, hence
the need for $\Subtype$.

@@

\emph{Notational conventions.} We name elements of $\Con$ as $\Gamma$, $\Delta$,
$\Theta$, elements of $\Sub\,\Gamma\,\Delta$ as $\sigma$, $\delta$, $\nu$,
elements of $\Ty\,\Gamma$ as $A$, $B$, $C$, and elements of $\Tm\,\Gamma\,A$ as
$t$, $u$, $v$. CwF components by default support de Bruijn indices, which are
not easily readable. We use instead a nameful notation for binders in context
extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon\,\ext (a : U) \ext (t :
\El\,a))$. We also define a type-theoretic flavor of $\appi$ for convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[\id, u])\\
  & t\appitt u && :\equiv (\appi\,t)[\id, u]
\end{alignat*}
We abbreviate non-dependent inductive $\Pii$ as $\blank\arri\blank$, and
likewise we use $\blank\arre\blank$ and $\blank\arrinf\blank$ for non-dependent
external and infinitary functions.

%% We use abbreviations for non-dependent function types:
%% \begin{alignat*}{3}
%%   & A\arri B && :\equiv \Pi\,A\,(B[\p])\\
%%   & A \arre\,B &&:\equiv \Pie\,A\,(\lambda\,\_.\,B) \\
%%   & A \arrinf\,B &&:\equiv \Piinf\,A\,(\lambda\,\_.\,B)
%% $A\arri B :\equiv \Pi\,A\,(B[\p])$, and $A \arre\,B :\equiv \Pie\,A\,(\lambda\,\_.\,B)$, and likewise
%% $A \arrinf\,B :\equiv \Piinf\,A\,(\lambda\,\_.\,B)$.

\begin{definition}[Notion of signature]\label{def:signature}
A QIIT signature at level $j$ is a context in an arbitrary $M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_j :\equiv (i : \mathbb{N})\ra(M : \ToS_{i,j})\ra \Con_M
\]

@@

\begin{example}{
    Signature for natural numbers. Here, no external types appear, so the level
    can be chosen as $0$.}
\begin{alignat*}{3}
  & \NatSig : \Sig_0 \\
  & \NatSig :\equiv \lambda (i :\mathbb{N})(M : \ToS_{i,0}).\\
  & \hspace{1em}(\emptycon_M\,\ext_M\, (N : \U_M) \,\ext_M\,(zero : \El_M\,N)\\
  & \hspace{2.67em}\ext_M (suc : N\arri_M\El_M\,N))
\end{alignat*}
\end{example}

@@

\begin{example}{Infinitary constructors}. The universe $\U$ is closed under
the $\Piinf$ function type, which allows such functions to appear in the domains
of $\Pii$ types. This allows, for example, a signature for trees branching with
arbitrary small types. This is a signature at level 1, since we have $\Set_0$ as
a $\Pie$ domain type.
\begin{alignat*}{3}
& \mathsf{TreeSig} :\equiv \\
& \hspace{1em}\emptycon \ext (Tree : \U)\\
& \hspace{1.35em}\ext (node : \Pie\,\Set_0\,(\lambda A.\,(A\,\arrinf\,Tree)\arri\El\,Tree))
\end{alignat*}
\end{example}

%% \begin{example}{Sort equations}. An interesting example is the term lifting
%% rule for cETT (omitting everything except term lifting):
%% \begin{alignat*}{3}
%%   & cETTSig :\equiv \\
%%   & ...\,\ext\,Tm{\Lift} : \Pii\,(\Gamma : Con)\,(\Pii\,(A : Ty{\appitt}\Gamma)((\IdU\,(Tm{\appitt}\Gamma{\appitt}A)\,(Tm{\appitt}\Gamma{\appitt}(\Lift_j{\appitt}A))))
%% \end{alignat*}
%% \end{example}

\begin{example}{Recursive equations}. Again, the universe is closed under $\Id$, which allows us to write equations
in $\Pii$ domains. A minimal (and trivial) example:
\begin{alignat*}{3}
  & \mathsf{RecEqSig} :\equiv\\
  & \hspace{0.5em}\emptycon\,\ext (A : \U) \ext (a : \El\,A)\ext (f : \Pii\,(x : A)\,(\Id\,A\,x\,a\arri\El\,A))
\end{alignat*}
More interesting (and complicated) examples for recursive equations are boundary
conditions in various cubical type theories \cite{cohen2016cubical, angiuli2016computational, angiuli2018cartesian}. Note that our $\Id$
allows iterated equations as well, but these are all trivial in the semantics,
where we assume UIP.
\end{example}

@@

\begin{definition}[Notion of induction in an flCwF]\label{def:induction}
Given $\bGamma : \flCwF_i$, we have the following predicate on contexts:
\begin{alignat*}{3}
  & \Inductive : \Con_{\bGamma} \ra \Set_i \\
  & \Inductive\,\Gamma :\equiv (A : \Ty_{\bGamma}\,\Gamma)\ra \Tm_{\bGamma}\,\Gamma\,A
\end{alignat*}
\end{definition}

@@

For $\boldsymbol{\sigma \circ \delta}$, the underlying functions are given by
function composition, and the preservation morphisms are given as follows:
\begin{alignat*}{3}
  & (\boldsymbol{\sigma \circ \delta})_{\emptycon}^{-1} :\equiv
    \bsigma\,\bdelta_{\emptycon}^{-1} \circ \bdelta_{\emptycon}^{-1} \\
  & (\boldsymbol{\sigma \circ \delta})_{\ext}^{-1} :\equiv
    \bsigma\,\bdelta_{\ext}^{-1} \circ \bdelta_{\ext}^{-1}
\end{alignat*}

@@

\subsection{Type Substitution}
We aim to define $\boldsymbol{\blank[\blank] : \Ty\,\Delta \ra
  \Sub\,\Gamma\,\Delta \ra \Ty\,\Gamma}$, such that $\boldsymbol{A[\id]} = \bA$
and $\boldsymbol{A[\sigma\circ\delta]} = \boldsymbol{A[\sigma][\delta]}$. The
underlying sets are given by simple composition:
\begin{alignat*}{3}
  & \Con_{\boldsymbol{A[\sigma]}}\,\ulGamma && :\equiv \Con_{\bA}\,(\bsigma\,\ulGamma)\\
  & \Sub_{\boldsymbol{A[\sigma]}}\,\Gamma\,\Delta\,\ulsigma && :\equiv
    \Sub_{\bA}\,\Gamma\,\Delta\,(\bsigma\,\ulsigma)\\
  & \Ty_{\boldsymbol{A[\sigma]}}\,\Gamma\,\ulA && :\equiv
      \Ty_{\bA}\,\Gamma\,(\bsigma\,\ulA)\\
  & \Tm_{\boldsymbol{A[\sigma]}}\,\Gamma\,A\,\ult && :\equiv
      \Tm_{\bA}\,\Gamma\,A\,(\bsigma\,\ult)
\end{alignat*}

Moreover, $\id_{\boldsymbol{A[\sigma]}} :\equiv \id_{\bA}$,
$\sigma\circ_{\boldsymbol{A[\sigma]}}\delta :\equiv \sigma\circ_{\bA}\delta$,
and likewise substitution components are given by corresponding substitution
components in $\bA$.

Context and type formers are given by coercing $\bA$
structures along $\bsigma$ preservation isomorphisms. For example:
\begin{alignat*}{3}
  &\emptycon_{\boldsymbol{A[\sigma]}} && :\equiv
    \coe\,\bsigma_{\emptycon}^{-1}\,\emptycon_{\bA}\\
  &\Gamma\ext_{\boldsymbol{A[\sigma]}}A && :\equiv
    \coe\,\bsigma_{\ext}^{-1}\,(\Gamma\ext_{\bA} A)\\
  &\Id_{\boldsymbol{A[\sigma]}}\,t\,u && :\equiv
    \coe\,\bsigma_{\Id}^{-1}\,(\Id_{\bA}\,t\,u)
\end{alignat*}

Term and substitution formers are given by composing $\coh$-lifted
isomorphisms with term and substitution formers from $\bA$. For example:
\begin{alignat*}{3}
  & \epsilon_{\boldsymbol{A[\sigma]}} && :\equiv
    \coh\,\bsigma_{\emptycon}^{-1}\,\emptycon_{\bA} \circ \epsilon_{\bA}\\
  & \p_{\boldsymbol{A[\sigma]}} && :\equiv
    \p_{\bA} \circ (\coh\,\bsigma_{\ext}^{-1}\,(\Gamma\ext A))^{-1}\\
  & \appK_{\boldsymbol{A[\sigma]}}\,t && :\equiv
    \appK_{\bA}\,((\coh\,\bsigma_{\K}\,(\K\,\Delta))^{-1}\circ t)
\end{alignat*}
Equations for term and type substitution follow from naturality of preservation
isomorphisms in $\bsigma$, $\coe[]$, $\coh[]$ and substitution equations in
$\bA$.

Iso-cleaving is given iso-cleaving in $\bA$ and the action of $\bsigma$ on
isomorphisms, e.g.\ we have $\coe_{\boldsymbol{A[\sigma]}}\,\ulsigma\,\Gamma
:\equiv \coe_{\bA}\,(\bsigma\,\ulsigma)\,\Gamma$.

@@

$\boldsymbol{\blank[\blank] : \Tm\,\Delta\,A \ra (\sigma : \Sub\,\Gamma\,\Delta)
  \ra \Tm\,\Gamma\,(A[\sigma])}$ is given similarly to
$\boldsymbol{\blank\circ\blank}$ in Section \ref{sec:idcomp}. Underlying functions
are given by function composition, and preservation morphisms are also similar:
\begin{alignat*}{3}
  & (\boldsymbol{t[\sigma]})_{\emptycon}^{-1} :\equiv
    \bt\,\bsigma_{\emptycon}^{-1} \circ \bt_{\emptycon}^{-1} \\
  & (\boldsymbol{t[\sigma]})_{\ext}^{-1} :\equiv
    \bt\,\bsigma_{\ext}^{-1} \circ \bt_{\ext}^{-1}
\end{alignat*}

@@

$\boldsymbol{\p : \Sub\,(\Gamma\ext A)\,\Gamma}$ is a strict morphism given by
taking a first projection for each component. $\boldsymbol{\q : \Tm\,(\Gamma\ext
  A)\,(A[\p])}$ is likewise a strict flCwF section given by second projections. Substitution extension $\boldsymbol{\sigma,\,t}$ is given by pointwise combining $\bsigma$ and $\bt$ with metatheoretic $\Sigma$ pairing, e.g.\ $\Con_{\boldsymbol{(\sigma,t)}}\,\ulGamma :\equiv (\bsigma\,\ulGamma,\,\bt\,\ulGamma)$.

@@

\begin{definition} For a level $i$, we write $\bSet_i$ for the flCwF of sets where $\Con_{\bSet_i} :\equiv \Set_i$ and $\Sub_{\bSet_i}\,\Gamma\,\Delta :\equiv \Gamma \ra \Delta$.
\end{definition}

We define $\boldsymbol{\U : \Ty\,\Gamma}$ as the isofibration which is
constantly $\bSet_i$. A constant isofibration does not actually depend on the
base flCwF, and has trivial iso-cleaving where $\coe$-s are identity
functions. Hence, we have $\Con_{\bU}\,\ulGamma :\equiv \Set_i$ and
$\Sub_{\bU}\,\Gamma\,\Delta\,\ulsigma :\equiv \Gamma \ra \Delta$.

@@

We define $\boldsymbol{\El : \Tm\,\Gamma\,\U \ra \Ty\,\Gamma}$ as discrete
isofibration formation. For $\boldsymbol{a : \Tm\,\Gamma\,\U}$, the underlying
sets of $\bEl\,\ba$ are the following:
\begin{alignat*}{3}
  & \Con_{\bEl\,\ba}\,\ulGamma && :\equiv \ba\,\ulGamma\\
  & \Sub_{\bEl\,\ba}\,\Gamma\,\Delta\,\ulsigma && :\equiv \ba\,\ulsigma\,\Gamma = \Delta\\
  & \Ty_{\bEl\,\ba}\,\Gamma\,\ulA && :\equiv \ba\,\ulA\,\Gamma\\
  & \Tm_{\bEl\,\ba}\,\Gamma\,A\,\ult && :\equiv \ba\,\ult\,\Gamma = A
\end{alignat*}
Hence, in $\bEl\,\ba$, $\Sub$ and $\Tm$ are propositional. We use the
isomorphisms $\ba_{\emptycon} : \ba\,\emptycon \simeq \top$ and $\ba_{\ext} :
\ba\,(\ulGamma\ext\ulA) \simeq (\Gamma :
\ba\,\ulGamma)\times(\ba\,\ulA\,\Gamma)$ to define empty context and context
extension:
\begin{alignat*}{3}
  & \emptycon_{\bEl\,\ba} && :\equiv \ba_{\emptycon}^{-1}\,\tt\\
  & (\Gamma\ext_{\bEl\,\ba} A) && :\equiv \ba_{\ext}^{-1}\,(\Gamma,\,A)
\end{alignat*}
We likewise use preservation isomorphisms to define $\K$, $\Id$ and $\Sigma$.
Context coercion is $\coe\,\ulsigma\,\Gamma :\equiv \ba\,\ulsigma\,\Gamma$. Type coercion, for $A : \ba\,\ulA\,\Gamma$ is given as $\coe\,\ult\,A :\equiv \ba\,\ult\,(\ba_{\ext}^{-1}\,(\Gamma,\,A))$.

@@

Contexts are products of $\bB$-contexts, and types are products of $\bB$-types,
indexed respectively by contexts and types of $\bEl\,\ba$.
\begin{alignat*}{3}
  & \Con_{(\bPii\,\ba\,\bB)}\,\ulGamma &&:\equiv (\gamma : \ba\,\ulGamma)\ra \Con_{\bB}\,(\ulGamma, \gamma)\\
  & \Ty_{(\bPii\,\ba\,\bB)}\,\Gamma\,\ulA &&:\equiv (\gamma : \ba\,\ulGamma)(a : \ba\,\ulA\,\gamma)\ra \Ty_{\bB}\,(\Gamma\,\gamma)\,(\ulA, a)
\end{alignat*}
Note that since $\bB$ is over the total $\boldsymbol{(\Gamma\ext\El\,a)}$,
$\Con_{\bB}$ has a $\Sigma$-typed argument, and likewise the last argument
of every $\bB$ component.  We could define substitutions similarly, as
products of substitutions:
\begin{alignat*}{3}
  & \Sub_{(\bPii\,\ba\,\bB)}\,\Gamma\,\Delta\,\ulsigma :\equiv
  (\gamma : \ba\,\ulGamma)(\delta : \ba\,\ulDelta)
  (\sigma : \Sub_{(\bEl\,\ba)}\,\gamma\,\delta\,\ulsigma)\\
  &\hspace{6.5em}\ra \Sub_{\bB}\,(\Gamma\,\gamma)\,(\Delta\,\delta)\,(\ulsigma, \sigma)
\end{alignat*}
This would work, but we know that $\Sub_{(\bEl\,\ba)}\,\gamma\,\delta\,\ulsigma$ is defined
as $\ba\,\ulsigma\,\gamma = \delta$, so we can eliminate $\sigma$ by singleton contraction,
and use the following equivalent definition:
\begin{alignat*}{3}
  & \Sub_{(\bPii\,\ba\,\bB)}\,\Gamma\,\Delta\,\ulsigma :\equiv
  (\gamma : \ba\,\ulGamma)\ra\Sub_{\bB}\,(\Gamma\,\gamma)\,(\Delta\,(\ba\,\ulsigma\,\gamma)\,(\ulsigma, \refl)
\end{alignat*}
The benefit of the contracted definition is that it computes preservation laws
in algebra homomorphisms strictly as expected, while the non-contracted
definition computes homomorphisms as functional logical relations.

@@

Iso-cleaving is given by transporting indices backwards in $\bEl\,\ba$ and outputs forwards
in $\bB$:
\begin{alignat*}{3}
  & \coe\,\ulsigma\,\Gamma &&:\equiv
    \lambda\,\gamma.\,\coe_{\bB}\,(\ulsigma,\refl)\,(\Gamma\,(\ba\,(\ulsigma^{-1})\,\gamma))\\
  & \coe\,\ult\,A &&:\equiv
    \lambda\,\gamma\,a.\,\coe_{\bB}\,(\ult,\refl)\,(A\,(\ba\,(\ult^{-1})\,(\ba_{\ext}^{-1}(\gamma,a))))
\end{alignat*}
Likewise, $\coh$-s are given by backwards-forwards $\coh$-s.

@@

The basic idea is that initial algebras can be built from the terms of $\syn_j$.
For example, consider the syntactic signature for natural numbers:
\[
\NatSig :\equiv \emptycon\ext(N : \U)\ext(zero : \El\,N)\ext (suc : N
\arri \El\,N)
\]
The type $\Tm_{\syn}\,\NatSig\,(\El_{\syn}\,N)$ is isomorphic to the usual type
of natural numbers, since, intuitively, such terms can only be built from
iterated usage of $zero$ and $suc$. We will build a term algebra for each
signature in this manner. First, we describe the interpretation of $\syn$ in the
$\Set$ model.
